---
title: "Weilin_Wang_Assignment3_DataExploration"
author: "Weilin Wang"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
# Load necessary packages
library(tidyverse)
library(lubridate)
library(here)

# Check current working directory
getwd()

# Load the datasets, reading strings as factors
Neonics <- read_csv(here("Data", "Raw", "ECOTOX_Neonicotinoids_Insects_raw.csv"), col_types = cols(.default = col_factor()))

Litter <- read_csv(here("Data", "Raw","NEON_NIWO_Litter_massdata_2018-08_raw.csv"), col_types = cols(.default = col_factor()))

# Confirm datasets have been loaded correctly
head(Neonics)
head(Litter)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are a class of systemic insecticides that affect the nervous systems of insects, leading to paralysis and death. Their use has been linked to Pollinator Decline. Pollinators like bees play a critical role in agriculture by pollinating crops, and their decline can have significant consequences for food production and biodiversity. The second effect could be water and Soil Contamination.These chemicals can persist in the environment, contaminating water bodies and soil, thus affecting aquatic insects and other invertebrates as well.

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Studying litter and woody debris in forests is crucial for understanding nutrient cycling, carbon storage, and ecosystem dynamics. As organic matter like leaves, twigs, and fallen branches decompose, they release nutrients back into the soil, which supports plant growth and maintains soil fertility. Additionally, woody debris plays a key role in carbon sequestration, acting as a temporary carbon sink and influencing the forest's capacity to mitigate climate change, while also providing habitat for various organisms and contributing to overall biodiversity.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.Litter Traps
 2.Sampling Frequency
 3. Woody Debris Transects

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
dim(Neonics)

dim(Litter)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}
effect_summary <- summary(Neonics$Effect)
sorted_effect_summary <- sort(effect_summary, decreasing = TRUE)

sorted_effect_summary

```

> Answer:These effects are of particular interest because they directly influence insect population dynamics, survival rates, and their ecological roles. Studying mortality helps assess the immediate lethality of neonicotinoids, while behavioral changes and reproductive impairment provide insights into sub-lethal, long-term impacts that may affect insect populations and ecosystem balance over time.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
summary(Neonics$`Species Common Name`, maxsum = 6)

```

> Answer:The six most commonly studied species might include honey bees, houseflies, mosquitoes, fruit flies, aphids, and beetles. These species are of interest because they either serve crucial ecological roles or are major agricultural pests.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
class(Neonics$`Conc 1 (Author).`)

head(Neonics$`Conc 1 (Author)`)
```

> Answer:column may not be numeric because it could contain non-numeric characters such as units or symbols like "<" indicating less than or greater than concentrations. These characters would prevent R from recognizing the column as numeric. 

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
# Load necessary library
library(ggplot2)

ggplot(Neonics, aes(x = as.numeric(`Publication Year`))) +
  geom_freqpoly(binwidth = 1, color = "blue") +
  labs(title = "Number of Studies by Publication Year",
       x = "Publication Year",
       y = "Number of Studies") +
  theme_minimal()

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
# Convert Publication Year to numeric and generate the plot with color by Test Location
ggplot(Neonics, aes(x = as.numeric(`Publication Year`), color = `Test Location`)) +
  geom_freqpoly(binwidth = 1) +
  labs(title = "Number of Studies by Publication Year (Colored by Test Location)",
       x = "Publication Year",
       y = "Number of Studies") +
  theme_minimal()


```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer:

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}


```

> Answer:

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}


```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}


```

> Answer:

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
