---
title: "WeilinWang_A07_GLMs.Rmd"
author: "Weilin Wang"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1

getwd()

library(tidyverse)
library(agricolae)
file_path <- "~/EDE_Fall2024/Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv"
lake_data <- read_csv(file_path)


#2
custom_theme <- theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(size = 10),
    legend.position = "right",
    legend.background = element_rect(fill = "lightgray", color = "darkgray"),
    legend.title = element_text(face = "bold"),
    panel.grid.major = element_line(color = "gray80"),
    panel.grid.minor = element_blank()
)

theme_set(custom_theme)

head(lake_data)

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: There's no difference in the mean lake temperature recorded during July across different depths in all lakes.
Ha: There is a difference in the mean lake temperature recorded during July across different depths in all lakes.


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
lake_data_july <- lake_data %>%
  mutate(sampledate = as.Date(sampledate, format = "%m/%d/%y")) %>%
  filter(format(sampledate, "%m") == "07") %>%
  select(lakename, year4, daynum, depth, temperature_C) %>%
  drop_na()

print(head(lake_data_july))

#5
ggplot(lake_data_july, aes(x = depth, y = temperature_C)) +
  geom_point(alpha = 0.6, color = "blue", size = 1.5) +
  geom_smooth(method = "lm", color = "red", se = FALSE, linetype = "dashed") +
  scale_y_continuous(limits = c(0, 35)) +
  labs(
    title = "Relationship Between Depth and Temperature in July Across Lakes",
    x = "Depth (m)",
    y = "Temperature (°C)"
  ) +
  custom_theme

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: It shows a clear negative relationship between depth and temperature in lakes during July, with temperature decreasing as depth increases. This suggests that deeper parts of the lake tend to be cooler, likely due to thermal stratification, where sunlight warms surface waters more than deeper layers. Although the trend appears mostly linear, the distribution of points indicates some non-linearity, especially at greater depths where temperature values spread out more. This spread suggests that the rate of temperature decrease may vary with depth, hinting at possible layers within the lake that impact temperature differently.


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression}
#7
model <- lm(temperature_C ~ depth, data = lake_data_july)

summary(model)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model results suggest a statistically significant relationship between lake depth and temperature in July. Specifically, 73.87% of the variability in temperature is explained by changes in depth, indicating that depth is a key factor influencing lake temperature. This model is based on 9,726 degrees of freedom, which provides a robust basis for the findings. The relationship is highly statistically significant, with a p-value less than 2e-16, far below typical significance thresholds. For every 1-meter increase in depth, the temperature is predicted to decrease about 1.95°C. This consistent decline in temperature with increasing depth highlights the cooling effect of deeper waters in the lake.


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
# Fit multiple models with different combinations of predictors
model1 <- lm(temperature_C ~ year4, data = lake_data_july)
model2 <- lm(temperature_C ~ daynum, data = lake_data_july)
model3 <- lm(temperature_C ~ depth, data = lake_data_july)
model4 <- lm(temperature_C ~ year4 + daynum, data = lake_data_july)
model5 <- lm(temperature_C ~ year4 + depth, data = lake_data_july)
model6 <- lm(temperature_C ~ daynum + depth, data = lake_data_july)
model7 <- lm(temperature_C ~ year4 + daynum + depth, data = lake_data_july)

# Compare models using AIC
AIC_results <- AIC(model1, model2, model3, model4, model5, model6, model7)
print(AIC_results)



#10
best_model <- lm(temperature_C ~ year4 + daynum + depth, data = lake_data_july)

summary(best_model)


```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The AIC method suggests that the best predictors in July are year4, daynum, and depth. This model explains 74.12% of the variance in temperature, a slight improvement over the depth-only model, which explained 73.87%. While depth remains the most significant factor, year and day of the year add minor but statistically significant explanatory power. Overall, including all three variables offers a marginally better prediction of lake temperature across this period.
---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
# ANOVA model
anova_model <- aov(temperature_C ~ lakename, data = lake_data_july)
summary(anova_model)
# Linear model
lm_model <- lm(temperature_C ~ lakename, data = lake_data_july)
summary(lm_model)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
I found there's a statistically significant difference in mean July temperatures among the lakes. The ANOVA results show an F-value of 50 with a p-val of < 2e-16, indicating strong evidence that temperatures vary by lake. Similarly, the linear model confirms that each lake’s average temperature significantly differs from the reference lake, with all lakes showing p-values below 0.05. However, the model’s adjusted R-squared value of 0.0387 suggests that lake identity explains only a small portion of the temperature variance.


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
library(ggplot2)

ggplot(lake_data_july, aes(x = depth, y = temperature_C, color = lakename)) +
  geom_point(alpha = 0.5, size = 1.5) + 
  geom_smooth(method = "lm", se = FALSE, linetype = "solid") + 
  scale_y_continuous(limits = c(0, 35)) + 
  labs(
    title = "Lake Temperature by Depth in July",
    x = "Depth (m)",
    y = "Temperature (°C)",
    color = "Lake Name"
  ) +
  custom_theme


```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
tukey_results <- TukeyHSD(anova_model)
print(tukey_results)

```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Peter Lake has a statistically similar mean temperature to Paul Lake and Ward Lake. East Long Lake stands out as it has a mean temperature significantly different from all other lakes.
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: I can be two-sample t-test. This test assesses whether there is a statistically significant difference between the means of two independent groups since we’re comparing temperatures between just these two lakes.


18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
lake_data_crampton_ward <- lake_data_july %>%
  filter(lakename %in% c("Crampton Lake", "Ward Lake"))

t_test_result <- t.test(temperature_C ~ lakename, data = lake_data_crampton_ward)
print(t_test_result)

```

>Answer: p-value: 0.2649 > 0.05. This suggests that we fail to reject the null hypothesis. And the 95% confidence interval for the difference in means ranges from -0.68 to 2.47, also supporting the conclusion that there’s no significant difference between the lakes' mean temperatures. Comparing to part 16, both tests confirm that their July temperatures are statistically similar.
